---
title: "DAG Analysis: Weather × Post-MoU Mortality Model"
author: "Giorgio Coppola"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    self-contained: true
editor:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 6, dpi = 150
)
```


## Research question

Did the 2017 EU-Libya MoU make the Central Mediterranean route more
dangerous (increase per-crossing mortality)?

**Identification strategy:** Weather × Post-MoU interaction ($\beta_3$)
captures how the MoU changed the Weather → vulnerability → Deaths pathway.

## 1. Define the DAG

Nodes and edges traced from `dag-1.pdf`:

| Node | Description | Status |
|------|-------------|--------|
| MoU | Feb 2017 EU-Libya policy shock | Exposure (treatment) |
| Deaths | Death count (per incident or aggregate) | Outcome |
| Vulnerability | Crossing vulnerability (boat quality, rescue, conditions, overcrowding) | **Latent (unobservable)** |
| Weather | Sea conditions: SWH, wind, SST | Observed, exogenous |
| FE | Seasonality + grid fixed effects | Adjusted (in model) |
| SAR | Search & rescue capacity | Mostly unobserved (low quality data) |
| Boats | Vessel type/quality | Partially observed |
| Crossings | Crossing volume | Observed, endogenous |

```{r dag-define}
library(dagitty)

dag <- dagitty('
dag {
  MoU       [exposure, pos="4,-0.5"]
  Deaths    [outcome,  pos="6.5,2"]
  Vulnerability      [latent,   pos="4,2"]
  Weather   [pos="0,2"]
  FE        [adjusted, pos="1.5,3.5"]
  SAR       [pos="3,0.5"]
  Boats     [pos="5,0.5"]
  Crossings [pos="1.5,1.1"]

  MoU -> SAR
  MoU -> Boats
  MoU -> Crossings
  SAR -> Crossings
  SAR -> Vulnerability
  Boats -> Crossings
  Boats -> Vulnerability
  Weather -> Crossings
  Weather -> Vulnerability
  Vulnerability -> Deaths
  Crossings -> Deaths
  FE -> Weather
  FE -> Crossings
  FE -> Vulnerability
  FE -> Deaths
}
')
```

```{r dag-plot, fig.height=5}
library(ggdag)
library(ggplot2)

ggdag_status(dag, text = FALSE) +
  geom_label(aes(label = name),
             fill = "grey93", colour = "black",
             size = 3, fontface = "bold",
             label.padding = unit(0.2, "lines"),
             label.size = 0, na.rm = TRUE) +
  theme_dag() +
  guides(color = guide_legend(title = "Node status")) +
  ggtitle("DAG: MoU → Deaths (Vulnerability unobserved)")
```

::: {.callout-note}
### Notes on edge directionality and omitted edges

**SAR ↔ Crossings (bidirectional in theory, unidirectional in the
DAG).** The DAG includes SAR → Crossings, representing the post-MoU
deterrence channel: reduced SAR capacity reduces crossings
(Rodriguez Sanchez et al. 2023, Model C; Hoffmann Pham & Komiyama
2024, ECM). However, the reverse — Crossings → SAR — is also
plausible, especially in the pre-MoU period: the migration crisis
drove the expansion of SAR operations (Mare Nostrum was launched in
response to the October 2013 Lampedusa shipwreck; NGO SAR expanded
as crossings surged). Rodriguez Sanchez et al. (2023) are explicit
that SAR is partly a consequence of crossing attempts, not just a
cause. In the post-MoU period that our model targets, the dominant
direction is SAR → Crossings (policy-driven SAR withdrawal reduces
crossings). The DAG represents this post-treatment causal direction.

**MoU → Crossings (direct deterrence channel).** The MoU reduced
crossings not only through SAR withdrawal but also through direct
deterrence mechanisms outside the SAR/Boats pathway: tightening of
Libya's southern border, detention of migrants in Libya, and
signaling effects that discouraged departure (Zambiasi & Albarosa
2025). This edge is included in the DAG for completeness. Since
MoU has no parents, it does not create backdoor paths.

**SAR → Boats (not included).** Deiana et al. (2024) show that
extensive SAR *enables* inflatable boat use — their entire
identification rests on SAR making unsafe boats viable. This
suggests SAR → Boats may exist as a causal edge. We do not include
it in the main DAG because (a) the MoU independently affected both
SAR and Boats through distinct policy channels (SAR externalization
vs. smuggler market disruption), and (b) adding SAR → Boats would
not change identification (all paths remain front-door from MoU,
which is parentless). However, the mechanism is real and should be
acknowledged: part of the boat-type shift documented by Deiana et
al. and Hoffmann Pham & Komiyama (2024) is mediated by SAR
availability rather than being a direct MoU effect.

**MoU → Vulnerability directly (not included).** The MoU shifted migrants
to more dangerous conditions in Libyan detention and exploitation
(documented by Zambiasi & Albarosa 2025). If pre-crossing
vulnerability feeds into at-sea crossing vulnerability, a direct
MoU → Vulnerability edge could exist. We omit it from the main DAG because
(a) Vulnerability as defined here captures *at-sea crossing vulnerability*
specifically, not pre-departure conditions, and (b) even if
included, it would not threaten identification since MoU has no
parents. This channel is relevant for the broader discussion of
MoU consequences but is outside the scope of the Weather × Post
identification strategy.
:::

## 2. All causal paths

### MoU → Deaths

```{r paths-mou}
p_mou <- paths(dag, from = "MoU", to = "Deaths")
data.frame(
  status = ifelse(p_mou$open, "OPEN", "closed"),
  path   = p_mou$paths
)
```

### Weather → Deaths

```{r paths-weather}
p_wx <- paths(dag, from = "Weather", to = "Deaths")
data.frame(
  status = ifelse(p_wx$open, "OPEN", "closed"),
  path   = p_wx$paths
)
```

## 3. Adjustment sets

### Total effect: MoU → Deaths

```{r adj-mou}
adj_mou <- adjustmentSets(dag, exposure = "MoU", outcome = "Deaths",
                           type = "minimal")
if (length(adj_mou) == 0) {
  cat("No adjustment needed — no backdoor paths from MoU to Deaths.\n")
  cat("MoU is exogenous: it has no parents in the DAG.")
} else {
  print(adj_mou)
}
```

MoU has no parents → no backdoor paths → the total effect is
identified without any conditioning.

### Total effect: Weather → Deaths

```{r adj-weather}
adjustmentSets(dag, exposure = "Weather", outcome = "Deaths",
               type = "minimal")
```

Must condition on **FE** to block the backdoor: Deaths ← FE → Weather.

### Canonical adjustment sets

```{r adj-canonical}
cat("MoU → Deaths:\n")
print(adjustmentSets(dag, exposure = "MoU", outcome = "Deaths",
                      type = "canonical"))

cat("\nWeather → Deaths:\n")
print(adjustmentSets(dag, exposure = "Weather", outcome = "Deaths",
                      type = "canonical"))
```

## 4. Checking specific conditioning strategies

```{r check-adj}
results <- data.frame(
  estimand = c(rep("MoU → Deaths", 3), rep("Weather → Deaths", 3)),
  conditioning_set = c(
    "{FE}", "{FE, Weather}", "{} (empty)",
    "{FE}", "{FE, MoU}", "{} (empty)"
  ),
  valid = c(
    isAdjustmentSet(dag, "MoU", "Deaths", c("FE")),
    isAdjustmentSet(dag, "MoU", "Deaths", c("FE", "Weather")),
    isAdjustmentSet(dag, "MoU", "Deaths", c()),
    isAdjustmentSet(dag, "Weather", "Deaths", c("FE")),
    isAdjustmentSet(dag, "Weather", "Deaths", c("FE", "MoU")),
    isAdjustmentSet(dag, "Weather", "Deaths", c())
  )
)
knitr::kable(results)
```

::: {.callout-note}
For the Weather × Post interaction model:

- The model conditions on **month FE** (= FE node) ✓
- Weather is exogenous to the MoU ✓
- The interaction $\beta_3$ captures the **change** in Weather's total
  effect on Deaths across periods
:::

## 5. Collider bias: conditioning on Crossings (Kronmal problem)

This is the DAG representation of **Kronmal (1993)**. When we use
mortality **rate** = Deaths/Crossings as the outcome, we implicitly
condition on Crossings.

Crossings is a **collider** between multiple parent nodes:

- Weather → Crossings ← SAR
- Weather → Crossings ← Boats

Conditioning on a collider **opens** spurious non-causal paths.

```{r collider-check}
cat("Is {FE, Crossings} valid for Weather → Deaths?\n")
cat(isAdjustmentSet(dag, "Weather", "Deaths", c("FE", "Crossings")))
```

### What paths open when we condition on Crossings?

```{r collider-paths}
dag_cond_cross <- dag
adjustedNodes(dag_cond_cross) <- c("FE", "Crossings")
p_cond <- paths(dag_cond_cross, from = "Weather", to = "Deaths")
data.frame(
  status = ifelse(p_cond$open, "OPEN", "closed"),
  path   = p_cond$paths
)
```

::: {.callout-warning}
**Kronmal problem in DAG terms:** conditioning on Crossings opens
collider paths through SAR and Boats:

- Weather → \[Crossings\] ← SAR → Vulnerability → Deaths
- Weather → \[Crossings\] ← Boats → Vulnerability → Deaths

The mortality **rate** inherits denominator-driven bias because
Crossings sits as a collider between Weather and SAR/Boats, which
also causally affect Deaths through Vulnerability.
:::

## 6. Can we fix collider bias by also conditioning on SAR + Boats?

```{r fix-collider}
cat("Is {FE, Crossings, SAR, Boats} valid for Weather → Deaths?\n")
cat(isAdjustmentSet(dag, "Weather", "Deaths",
                    c("FE", "Crossings", "SAR", "Boats")))
```

```{r fix-collider-paths}
dag_cond_fix <- dag
adjustedNodes(dag_cond_fix) <- c("FE", "Crossings", "SAR", "Boats")
p_fix <- paths(dag_cond_fix, from = "Weather", to = "Deaths")
data.frame(
  status = ifelse(p_fix$open, "OPEN", "closed"),
  path   = p_fix$paths
)
```

::: {.callout-note}
If SAR and Boats were fully observed, conditioning on
{FE, Crossings, SAR, Boats} would block the collider paths and
allow using the mortality rate. But SAR and Boats are only
**partially observed** (and Vulnerability remains latent), so this fix is
not practically available.
:::

## 7. Implied conditional independencies

These are **testable** predictions of the DAG. If any of these fail
empirically, the DAG structure may be wrong.

```{r cond-indep}
ci <- impliedConditionalIndependencies(dag)
for (s in ci) cat(format(s), "\n")
```

## 8. Instrumental variables

```{r iv}
cat("For MoU → Deaths:\n")
iv_mou <- instrumentalVariables(dag, exposure = "MoU", outcome = "Deaths")
if (length(iv_mou) == 0) {
  cat("  None needed (MoU is exogenous, no backdoor paths)\n")
} else {
  print(iv_mou)
}

cat("\nFor Weather → Deaths:\n")
iv_wx <- instrumentalVariables(dag, exposure = "Weather",
                                outcome = "Deaths")
if (length(iv_wx) == 0) {
  cat("  None found\n")
} else {
  print(iv_wx)
}
```

## 9. Sensitivity: adding Crossings → Vulnerability

Overcrowding from high volume could increase vulnerability. This adds
a new causal chain: Weather → Crossings → Vulnerability → Deaths. Since FE →
SAR and FE → Boats are absorbed by conditioning on FE, we only test
this edge.

```{r sensitivity-crossings-Vulnerability}
dag_ext <- dagitty('
dag {
  MoU       [exposure, pos="4.5,-0.5"]
  Deaths    [outcome,  pos="6.5,2"]
  Vulnerability      [latent,   pos="4,2"]
  Weather   [pos="0,2"]
  FE        [adjusted, pos="2.5,4"]
  SAR       [pos="2.5,0.5"]
  Boats     [pos="5,0.5"]
  Crossings [pos="0.5,3"]

  MoU -> SAR
  MoU -> Boats
  MoU -> Crossings
  SAR -> Crossings
  SAR -> Vulnerability
  Boats -> Crossings
  Boats -> Vulnerability
  Weather -> Crossings
  Weather -> Vulnerability
  Vulnerability -> Deaths
  Crossings -> Deaths
  Crossings -> Vulnerability
  FE -> Weather
  FE -> Crossings
  FE -> Vulnerability
  FE -> Deaths
}
')

cat("With Crossings → Vulnerability added:\n\n")

cat("Adjustment sets for Weather → Deaths:\n")
print(adjustmentSets(dag_ext, exposure = "Weather", outcome = "Deaths",
                      type = "minimal"))

cat("\nAdjustment sets for MoU → Deaths:\n")
adj_ext <- adjustmentSets(dag_ext, exposure = "MoU", outcome = "Deaths",
                           type = "minimal")
if (length(adj_ext) == 0) {
  cat("  Still no adjustment needed (MoU still exogenous)\n")
} else {
  print(adj_ext)
}
```

```{r sensitivity-collider-worse}
cat("Is {FE, Crossings} still invalid for Weather → Deaths?\n")
cat(isAdjustmentSet(dag_ext, "Weather", "Deaths", c("FE", "Crossings")))

cat("\n\nPaths Weather → Deaths | {FE, Crossings} (extended DAG):\n")
dag_ext_cond <- dag_ext
adjustedNodes(dag_ext_cond) <- c("FE", "Crossings")
p_ext <- paths(dag_ext_cond, from = "Weather", to = "Deaths")
data.frame(
  status = ifelse(p_ext$open, "OPEN", "closed"),
  path   = p_ext$paths
)
```

::: {.callout-important}
Adding Crossings → Vulnerability does **not** threaten identification (still
a front-door causal path, no new backdoor). But it makes the
**Kronmal problem worse**: conditioning on Crossings now also blocks
the real causal path Weather → Crossings → Vulnerability → Deaths, on top of
opening the collider bias paths. The mortality rate becomes an even
more distorted estimand.
:::

## 10. Counterfactual: what if Vulnerability were observed?

```{r Vulnerability-observed}
dag_obs <- dag
latents(dag_obs) <- c()

cat("Minimal adjustment sets for Weather → Deaths (Vulnerability observed):\n")
print(adjustmentSets(dag_obs, exposure = "Weather", outcome = "Deaths",
                      type = "minimal"))

cat("\nMinimal adjustment sets for MoU → Deaths (Vulnerability observed):\n")
print(adjustmentSets(dag_obs, exposure = "MoU", outcome = "Deaths",
                      type = "minimal"))
```

If Vulnerability were observed:

- Conditioning on {FE, Vulnerability} would block the Weather → Vulnerability → Deaths
  front-door path and **isolate the volume channel** only
- NOT conditioning on Vulnerability gives the **total effect** (both channels)
- This highlights that our Weather × Post interaction estimates the
  total effect change (danger + volume), not the pure danger channel

## 11. Estimand decomposition and model hierarchy

Per-crossing mortality can be decomposed as:

$$\frac{\text{deaths}_t}{\text{crossings}_t} = \underbrace{\frac{\text{incidents}_t}{\text{crossings}_t}}_{\text{frequency}} \times \underbrace{\frac{\text{deaths}_t}{\text{incidents}_t}}_{\text{severity}}$$

This is the **preferred theoretical estimand**
$E[\text{deaths} \mid \text{crossing}]$: it isolates per-crossing
danger from volume effects. However, it is **infeasible** as a
primary analysis for two reasons:

1. **Crossings data are monthly only** (interception data from
   LCG/TCG is monthly), so the per-crossing rate cannot be computed
   at event level.
2. **Crossings is a post-treatment variable** affected by MoU. Per
   the DAG (Section 5), conditioning on Crossings opens collider
   paths through SAR and Boats. This is only fixable by also
   conditioning on SAR and Boats (Section 6), which are not fully
   observed (Boats data requested from Frontex; SAR data partially
   available).

### Alternative: monthly count decomposition

An alternative decomposition avoids the crossings denominator:

$$\text{deaths}_t = \underbrace{\text{incidents}_t}_{\text{frequency}} \times \underbrace{E[\text{deaths} \mid \text{incident}]_t}_{\text{severity}}$$

This is also an accounting identity. It does **not** require
crossings — no collider opens. However, the frequency term
($\text{incidents}_t$) is a count, not a probability per crossing.
It conflates two distinct channels:

- **Volume**: more crossings → more incidents (regardless of
  per-crossing danger)
- **Per-crossing danger**: each crossing more likely to become an
  incident

Without crossings data in the denominator, these two channels
cannot be separated within a single model.

### Model hierarchy

| Model | Estimand | Data level | Crossings needed? | What $\beta_3$ captures |
|-------|----------|------------|-------------------|------------------------|
| **A** (primary) | $E[\text{deaths} \mid \text{incident}, \text{weather}, \text{post}]$ | Event-level | No | Did weather kill more people **per incident** after MoU? (severity margin) |
| **Monthly death counts** | $E[\text{deaths}_t \mid \text{weather}, \text{post}]$ | Monthly | No | Did more people **die in total** in bad-weather months after MoU? Mixes volume + danger |
| **Crossings diagnostic** | $E[\text{crossings}_t \mid \text{weather}, \text{post}]$ | Monthly | As outcome, not conditioning | Did weather's effect on **crossing volume** change after MoU? (volume channel assessment; see Section 12) |
| **D** (exploratory rate) | $\text{deaths}_t / \text{crossings}_t \sim \text{weather} \times \text{post}$ | Monthly | Yes → collider bias | Per-crossing danger, but collider bias from SAR/Boats (see Section 13) |
| **C** (aspirational) | As D, conditioning on {FE, Crossings, SAR, Boats} | Monthly | Yes, with SAR + Boats fix | Per-crossing danger, collider-free — requires SAR and Boats data |

**Model A** is the primary analysis: it captures severity at event
level without conditioning on crossings or incident occurrence
beyond the detection selection inherent in IOM MMP data. The monthly
death count model is complementary (captures the total effect). The
crossings diagnostic (Section 12) helps disentangle volume from
danger. Model D is exploratory with acknowledged collider bias.
Model C is theoretically preferred but requires data not yet
available.

## 12. Volume channel diagnostic: Crossings ~ Weather × Post

To partially disentangle the volume and danger channels in the
monthly death count model, we can model crossings as an **outcome**
in a separate regression:

$$\text{crossings}_t \sim f(\text{weather}_t \times \text{Post}_t + \text{month FE})$$

This models crossings as an **outcome**, not as a conditioning
variable — so no collider problem arises.

### DAG validity

```{r adj-weather-crossings}
cat("Adjustment sets for Weather → Crossings:\n")
print(adjustmentSets(dag, exposure = "Weather",
                      outcome = "Crossings",
                      type = "minimal"))

cat("\nIs {FE} valid for Weather → Crossings?\n")
cat(isAdjustmentSet(dag, "Weather", "Crossings", c("FE")))
```

```{r paths-weather-crossings}
p_wx_cross <- paths(dag, from = "Weather", to = "Crossings")
data.frame(
  status = ifelse(p_wx_cross$open, "OPEN", "closed"),
  path   = p_wx_cross$paths
)
```

The only paths from Weather to Crossings are:

- Weather → Crossings (direct causal edge — the one we want)
- Weather ← FE → Crossings (backdoor through FE — blocked by
  conditioning on FE)

No other paths exist because: (a) Weather's only parent in the DAG
is FE, and (b) the path Weather → Vulnerability → Deaths ← Crossings has a
collider at Deaths, which is blocked unless we condition on Deaths
(we do not in this regression).

Conditioning on **{FE}** is sufficient. The regression is **valid
per the DAG**.

::: {.callout-note}
Crossings is only a problem when it appears in the **conditioning
set** of a regression targeting Weather → Deaths (collider). When
Crossings is the **outcome**, there is no collider issue — we are
modelling it, not conditioning on it. SAR and Boats do not need to
be included because they do not cause Weather (they are not
confounders of the Weather → Crossings relationship).
:::

### Interpretation

$\beta_3$ in this regression captures whether the MoU changed how
weather affects crossing volume. Pre-MoU, bad weather strongly
deters crossings because SAR makes waiting costless (Deiana et al.
2024). Post-MoU, smugglers may be less responsive to weather
conditions.

This serves as a **diagnostic**: if weather's effect on crossings
did not change much post-MoU, then $\beta_3$ in the monthly death
count model is primarily about the **danger channel**
(Weather → Vulnerability → Deaths), not the volume channel
(Weather → Crossings → Deaths). This is a mediation-style argument
without formal mediation analysis, but it is informative and avoids
the collider problem entirely.

## 13. Rate regression as exploratory analysis (Model D)

The theoretical estimand
$E[\text{deaths} \mid \text{crossing}]$ is the most substantively
meaningful measure of route danger. It can be operationalized as a
rate regression:

$$\log(\text{death rate}_t + c) \sim \beta_1 \text{weather}_t + \beta_2 \text{Post}_t + \beta_3 (\text{weather}_t \times \text{Post}_t) + \lambda_m$$

or equivalently as a count model with offset:

$$\text{deaths}_t \sim \text{NegBin}\!\bigl(\text{weather} \times \text{Post} + \text{offset}(\log \text{crossings}_t) + \text{month FE}\bigr)$$

Both formulations condition on Crossings, which opens the collider
paths documented in Section 5.

### Why $\beta_3$ is partially protected by differencing

The Kronmal-induced bias from the denominator exists in **both**
the pre- and post-MoU periods (the Weather → Crossings relationship
is present throughout). The interaction $\beta_3$ captures the
**change** in the weather–mortality relationship across periods. If
the denominator-driven bias were constant across periods, it would
difference out.

Moreover, the death rate has clear substantive meaning: fewer
crossings with equal deaths implies higher per-crossing danger —
which is the real phenomenon.

### Why this argument is incomplete

The collider bias is **not** constant across periods. The spurious
paths opened by conditioning on Crossings are:

- Weather → \[Crossings\] ← **SAR** → Vulnerability → Deaths
- Weather → \[Crossings\] ← **Boats** → Vulnerability → Deaths

The MoU changed both SAR (withdrawal) and Boats (shift to
inflatables). Pre-MoU, SAR is high and Boats are predominantly
wooden → one collider bias magnitude. Post-MoU, SAR is low and
Boats are inflatable → a different magnitude. Therefore, $\beta_3$
picks up both the **true change** in weather-lethality and the
**change in collider bias** across periods. The direction of the
additional bias is ambiguous without knowing the exact structural
parameters.

### How to use Model D

Model D remains informative as an **exploratory/robustness**
analysis:

- Run with acknowledged collider bias
- Compare $\beta_3$ with and without crossings in the denominator
  (i.e., compare with the monthly death count model) to assess the
  empirical magnitude of the collider distortion
- If Frontex boat-type data become available, add Boats as a
  control (partially closing the collider)
- If SAR proxy data are available (e.g., NGO fleet presence), add
  SAR as well (approaching the Model C ideal of Section 6)
- Report alongside Model A (primary) and monthly counts
  (complementary) to show robustness

::: {.callout-important}
Model D is **not** the primary analysis. It is exploratory evidence
that the per-crossing danger signal exists, reported with
transparent discussion of the collider bias and its likely
direction.
:::

## 14. Model A: specification notes

### Why negative binomial?

Deaths per incident is a non-negative count with clear overdispersion:
most incidents produce a few deaths, but mass-casualty shipwrecks can
produce hundreds. The IOM MMP data show ~23% zero-death incidents on
the CMR (490/2,090), with a long right tail. The Poisson assumption
(mean = variance) is almost certainly violated.

The negative binomial adds a dispersion parameter $\alpha$ that
allows variance $> $ mean. Standard diagnostics to justify this
choice:

1. **Likelihood ratio test**: NegBin vs. Poisson (H0: $\alpha = 0$).
   If rejected, overdispersion is confirmed.
2. **Cameron–Trivedi test** for overdispersion in the Poisson model.
3. **AIC/BIC comparison** between Poisson and NegBin fits.

The Poisson specification should be reported as a **robustness check**.
Deiana et al. (2024) use Poisson QMLE for their crossing counts,
which is consistent under heteroskedasticity even if the variance
function is misspecified. If $\hat{\beta}_3$ is similar across
NegBin, Poisson, and DML (pipeline step 2), the functional form
choice is not driving results.

**Alternatives considered:**

| Model | Issue for this setting |
|-------|----------------------|
| Poisson | Mean = variance almost certainly violated (mix of small rubber-boat incidents and mass-casualty shipwrecks) |
| Quasi-Poisson | Robust to overdispersion but less efficient; does not model the dispersion structure |
| Zero-inflated NB | Defensible given 23% zeros, but zeros likely reflect reporting (missing confirmed dead, not a separate DGP) |
| OLS on $\log(\text{deaths} + 1)$ | Arbitrary constant, loses count structure, inconsistent under heteroskedasticity |

### Weather covariate selection and multicollinearity

**Selection rationale (theory-driven, not data-driven).** The 38
covariates are retained if and only if they satisfy two criteria:

1. **Exogenous to migration policy** — atmospheric/oceanic conditions
   cannot be affected by the MoU.
2. **Physically related to crossing danger** — waves capsize boats,
   wind drives drift, SST affects hypothermia risk, currents affect
   trajectory.

This gives: waves (SWH), wind speed, SST, current speed, cloud cover,
precipitation — each at lag 0 (conditions at time of incident) and
lag 1 (conditions day before, affecting departure decisions) = 19
variables $\times$ 2 lags = 38.

**Multicollinearity.** Weather variables are highly correlated
($r = 0.58$–$0.89$). This matters differently across the two models
in the analysis:

- **BSTS diagnostic (Section 11 of presentation):** all 38 enter as
  predictors. Spike-and-slab regularization handles multicollinearity
  by construction — it selects among correlated predictors. The
  finding that none are retained for the mortality rate (0/49 at
  $>$10% inclusion) is the diagnostic result.
- **NegBin Model A:** the triple interaction uses **SWH only** (or
  PC1 from PCA on all weather features). Other weather variables
  enter $\mathbf{X}_i$ as controls. Multicollinearity among weather
  controls does not bias $\hat{\beta}_3$ — it only affects the
  precision of the individual weather coefficients in $\hat{\gamma}$,
  which are not parameters of interest.

**PCA option:** Collapse the correlated weather dimensions into PC1
("overall sea harshness") for the triple interaction. This avoids
multicollinearity entirely but sacrifices individual-variable
interpretability. Both approaches (SWH alone, PC1) should be
reported.

### Interpreting $\beta_3$

In the NegBin framework with log link:

$$\log \mu_i = \alpha_g + \lambda_m + \beta_3 (\text{SWH}_i \times
\text{Post}_t \times \text{CMR}_r) + \mathbf{X}_i'\gamma$$

$\beta_3$ is a **triple-difference semi-elasticity**:

> For each additional 1m of significant wave height, the expected
> number of deaths per incident on the CMR post-MoU is
> $\exp(\beta_3)$ times what it would be based on the pre-MoU
> weather–mortality gradient (or the gradient on control routes).

**Substantive example:** if $\hat{\beta}_3 = 0.3$, then
$\exp(0.3) \approx 1.35$, meaning a 1m SWH increase raises expected
deaths per incident by 35% **more** on the CMR post-MoU than the
same increase did pre-MoU or on control routes. If the pre-MoU
gradient was essentially zero (SAR rescued distressed boats regardless
of weather), then $\beta_3$ captures nearly the full post-MoU
weather-lethality effect.

**Scaling note (Deiana et al. 2024):** small changes in SWH translate
to large changes in maximum wave height encountered. The Rayleigh
distribution implies max wave $\approx 2 \times$ SWH. A modest
$\beta_3$ per meter of SWH can therefore correspond to substantively
large lethality differences in the waves actually experienced.
Marginal effects should be reported at meaningful SWH values (median,
75th percentile) rather than only the raw coefficient.

## 15. Adjustment set visualization

```{r dag-adj-plot, fig.height=5}
ggdag_adjustment_set(dag, text = FALSE) +
  geom_label(aes(label = name),
             fill = "grey93", colour = "black",
             size = 3, fontface = "bold",
             label.padding = unit(0.2, "lines"),
             label.size = 0, na.rm = TRUE) +
  theme_dag() +
  ggtitle("Required adjustments: Weather → Deaths")
```

## 16. Summary

| Finding | Implication |
|---------|-------------|
| **MoU → Deaths**: no backdoor paths | Total effect identified without conditioning (MoU is exogenous) |
| **Weather → Deaths**: one backdoor through FE | Conditioning on FE (month + grid FE) is **sufficient** |
| **Conditioning on Crossings** (= using rate) | Opens collider paths through SAR/Boats → **Kronmal bias** |
| **Vulnerability is latent** | Weather × Post interaction identifies MoU's effect on the Weather → Vulnerability → Deaths pathway **without** observing Vulnerability |
| **Exclusion restriction** | No direct Weather × Post → Deaths path outside Vulnerability must hold |
| **Per-crossing decomposition infeasible** | Crossings only monthly + post-treatment → collider. Model A (event-level severity) is primary; monthly counts complementary |
| **Crossings ~ Weather × Post is valid** | {FE} sufficient; no collider because Crossings is outcome, not conditioning set. Diagnostic for volume channel |
| **Rate regression (Model D)** | $\beta_3$ partially protected by differencing, but collider bias changes across periods (SAR/Boats shift). Exploratory only |

### Data needs

| Variable | Status | Source |
|----------|--------|--------|
| Weather (SWH, wind, SST) | **Observed** | ERA5 |
| Deaths | **Observed** | IOM MMP |
| FE (date, grid) | **Observed** | Date/coordinates |
| SAR | Mostly unobserved (low quality data) | NGO fleet data |
| Boats | Partially observed | Frontex JORA |
| Crossings | Observed — **do not condition on** in primary model; use as **outcome** in diagnostic (Section 12); use as **denominator** only in exploratory Model D (Section 13) | IOM/UNHCR monthly |
| Vulnerability | **Unobserved** | No single data source |

### Potential missing edges

| Edge | Plausible? | Impact on identification |
|------|-----------|------------------------|
| FE → SAR | Yes (seasonal SAR deployment) | **None** — FE is already conditioned on, so any path through FE is blocked |
| FE → Boats | Yes (seasonal boat availability) | **None** — same reason, absorbed by FE conditioning |
| Crossings → Vulnerability | Yes (overcrowding from high volume) | Adds a new **causal** front-door path (Weather → Crossings → Vulnerability → Deaths) but does not create backdoor bias. Makes conditioning on Crossings (= using the rate) **even more problematic**: it would block this additional real causal variation. Tested in Section 9 |
| Crossings → SAR | Yes in pre-MoU period (migration crisis drove SAR expansion) | Not included because the DAG represents the post-MoU causal structure. In this period, the dominant direction is SAR → Crossings (policy-driven deterrence). See discussion in Section 1 callout |
| SAR → Boats | Yes (Deiana et al. 2024: SAR enables inflatable boat use) | **None** — all paths remain front-door from MoU (parentless). Not included but acknowledged as a real mechanism. See discussion in Section 1 callout |
| MoU → Vulnerability directly | Possible (detention conditions, pre-departure exploitation) | **None** — MoU has no parents. Not included because Vulnerability here captures at-sea vulnerability specifically. Relevant for broader discussion of MoU consequences |
| Weather → Deaths directly | No — every mechanism (waves capsizing boats, storms preventing rescue, exposure at sea) operates through crossing vulnerability | Not applicable; exclusion restriction is defensible |

### Literature support for DAG edges

| Edge | Supporting literature |
|------|---------------------|
| MoU → SAR | Zambiasi & Albarosa (2025): SAR externalized to LCG; Deiana et al. (2024): Minniti Code withdrew NGO SAR; Rodriguez Sanchez et al. (2023): EU-Libya cooperation = "shift to deterrence"; Hoffmann Pham & Komiyama (2024): rescue probability falls from ~90% to ~50%; Steinhilper & Gruijters (2018): mortality rates rose when SAR scaled back |
| MoU → Boats | Deiana et al. (2024): inflatable share drops post-Minniti; Hoffmann Pham & Komiyama (2024): smugglers shift to smaller boats as interception rises; Friebel et al. (2024): migration choices sensitive to crossing costs, including boat availability |
| MoU → Crossings | Hoffmann Pham & Komiyama (2024): deterrence via rescue probability; Zambiasi & Albarosa (2025): southern border tightening, detention deterrence; Rodriguez Sanchez et al. (2023): Model C shows pushback period reduces crossings; Friebel et al. (2024): elasticity of migration intentions to smuggling distance > −3; Tafani & Riccaboni (2025): EU-Turkey deal diverted ~2,000 migrants to CMR |
| SAR → Crossings | Hoffmann Pham & Komiyama (2024): ECM, rescue probability fall 0.9→0.5 predicts ~11,200 fewer monthly crossings; Rodriguez Sanchez et al. (2023): null at aggregate monthly level but significant for pushback period; Battiston (2022): reducing rescue distance increases future departures; Cusumano & Villa (2019): no pull factor from NGO SAR; Amenta et al. (2021): SAR not a pull factor |
| SAR → Vulnerability | Deiana et al. (2024): SAR withdrawal = no rescue for distressed boats; Zambiasi & Albarosa (2025): LCG's poor SAR increases deadly event probability by 50% per 100km from Libya; Battiston (2022): rescue distance directly affects safety; Fasani et al. (2019): SAR operations related to probability of dying; Steinhilper & Gruijters (2018): mortality rates rose when SAR scaled back |
| Boats → Vulnerability | Deiana et al. (2024): inflatables ~20× more dangerous than wooden boats (θ ≈ 5-10%); Hoffmann Pham & Komiyama (2024): boat size affects survival |
| Weather → Crossings | Deiana et al. (2024): 0.10m SWH increase reduces crossings 8.9-14.6%; Camarena et al. (2020): 10% wave height increase → ~27% fewer arrivals; Cusumano & Villa (2019): weather conditions are the main determinant of departures from Libya |
| Weather → Vulnerability | Deiana et al. (2024): weather × boat-type interaction — weather more dangerous for fragile boats; Camarena et al. (2020): rougher seas raise death risk (daily level). Empirical link at monthly frequency not yet fully explored in our analysis |
| Vulnerability → Deaths | All papers: definitional — vulnerability determines whether crossings result in death. Steinhilper & Gruijters (2018): descriptive analysis of mortality risk 2010-2016 |
| Crossings → Deaths | Tafani & Riccaboni (2025): crossings coeff = 0.001 (p=0.012) on missing migrants; Amenta et al. (2021): crossing volume as determinant of deaths |
| FE → {all} | Universal: every paper conditions on time/seasonal FE. Deiana et al. (2024): week-by-year FE; Camarena et al. (2020): month + day-of-week FE; Zambiasi & Albarosa (2025): month + quadrant FE + quadrant-specific trends |

### Structural assumptions: literature support

| Assumption | Supporting literature |
|-----------|---------------------|
| MoU exogeneity (no parents) | Zambiasi & Albarosa (2025): geopolitical agreement endorsed at EU Malta summit; Hoffmann Pham & Komiyama (2024): treats LCG shift as exogenous policy change |
| Weather ⊥ MoU | By physics: atmospheric/oceanic conditions are independent of political agreements |
| Exclusion restriction (no Weather → Deaths bypassing Vulnerability) | Deiana et al. (2024): weather's effect on danger is mediated by boat vulnerability; Camarena et al. (2020): all identified mechanisms (capsizing, failed rescue, exposure) operate through crossing vulnerability |
| Vulnerability is latent | No paper fully observes crossing vulnerability. Deiana et al. (2024): observe boat type only; Hoffmann Pham & Komiyama (2024): observe boat size and passengers only; Zambiasi & Albarosa (2025): observe location relative to SAR zones only |
| Collider bias from conditioning on Crossings | Kronmal (1993): foundational analysis of ratio-outcome bias; Elwert & Winship (2014): formalization of collider bias in DAG terms; Camarena et al. (2020): their death-rate regression is an example of this problem |
| Weather × Post interaction as identification | Novel in this context. Closest precedent: Deiana et al. (2024) use weather × SAR × boat-type triple interaction for crossing behavior, not mortality directly |

## References

1. Amenta, C., Di Betta, P., & Ferrara, C. (2021). The migrant
   crisis in the Mediterranean Sea: Empirical evidence on policy
   interventions. *Socio-Economic Planning Sciences*, 75, 100880.

2. Battiston, G. (2022). Rescue on stage: Border enforcement and
   public attention in the Mediterranean Sea. Marco Fanno Working
   Paper No. 0292.

3. Camarena, K. R., Claudy, S., Wang, Y., & Wright, A. L. (2020).
   Political and environmental risks influence migration and human
   smuggling across the Mediterranean Sea. *SSRN Working Paper*.

4. Cusumano, E., & Villa, M. (2019). Sea rescue NGOs: A pull
   factor of irregular migration? *EUI Policy Brief*.

5. Deiana, C., Maheshri, V., & Mastrobuoni, G. (2024). Migrants
   at sea: Unintended consequences of search and rescue
   operations. *American Economic Journal: Economic Policy*.

6. Elwert, F., & Winship, C. (2014). Endogenous selection bias:
   The problem of conditioning on a collider variable. *Annual
   Review of Sociology*, 40, 31--53.

7. Fasani, F., Frattini, T., & Weisser, R. (2019). The
   relationship between SAR operations and the probability of
   dying in the Central Mediterranean, 2012--2017. *Working
   Paper*.

8. Friebel, G., Manchin, M., Mendola, M., & Prarolo, G. (2024).
   International migration and illegal costs: Evidence from
   Africa-to-Europe smuggling routes. *Journal of International
   Economics*, 148, 103876.

9. Hoffmann Pham, K., & Komiyama, Y. (2024). Strategic choices of
   migrants and smugglers in the Central Mediterranean Sea.
   *Working Paper*.

10. Kronmal, R. A. (1993). Spurious correlation and the fallacy
    of the ratio standard revisited. *Journal of the Royal
    Statistical Society: Series A*, 156(3), 379--392.

11. Rodriguez Sanchez, A., Wucherpfennig, J., Rischke, R., &
    Iacus, S. M. (2023). Search-and-rescue in the Central
    Mediterranean Route does not induce migration: Predictive
    modeling to answer causal queries in migration research.
    *Scientific Reports*, 13, 12871.

12. Steinhilper, E., & Gruijters, R. J. (2018). A contested
    crisis: Policy narratives and empirical evidence on border
    deaths in the Mediterranean. *Sociology*, 52(3), 515--533.

13. Tafani, S., & Riccaboni, M. (2025). The impact of the
    EU--Turkey agreement on the number of lives lost at sea.
    *Humanities and Social Sciences Communications*, 12, 282.

14. Textor, J., van der Zander, B., Gilthorpe, M. S., Liskiewicz,
    M., & Ellison, G. T. (2016). Robust causal inference using
    directed acyclic graphs: The R package 'dagitty'.
    *International Journal of Epidemiology*, 45(6), 1887--1894.

15. Zambiasi, D., & Albarosa, E. (2025). Externalizing rescue
    operations at sea: The migration deal between Italy and Libya.
    *Journal of Economic Geography*, 25(1), 41--58.
