---
title: "DAG Analysis: Weather × Post-MoU Mortality Model"
author: "Giorgio Coppola"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    self-contained: true
editor:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 6, dpi = 150
)
```


## Research question

Did the 2017 EU-Libya MoU make the Central Mediterranean route more
dangerous (increase per-crossing mortality)?

**Identification strategy:** Weather × Post-MoU interaction ($\beta_3$)
captures how the MoU changed the Weather → vulnerability → Deaths pathway.

## 1. Define the DAG

Nodes and edges traced from `dag-1.pdf`:

| Node | Description | Status |
|------|-------------|--------|
| MoU | Feb 2017 EU-Libya policy shock | Exposure (treatment) |
| Deaths | Death count (per incident or aggregate) | Outcome |
| Vulnerability | Crossing vulnerability (boat quality, rescue, conditions, overcrowding) | **Latent (unobservable)** |
| Weather | Sea conditions: SWH, wind, SST | Observed, exogenous |
| FE | Time/seasonality FE (week-year in daily panel; grid + yr + mo in event model) | Adjusted (in model) |
| SAR | Search & rescue capacity | Mostly unobserved (low quality data) |
| Boats | Vessel type/quality | Partially observed |
| Crossings | Crossing volume | Observed, endogenous |

```{r dag-define}
library(dagitty)

dag <- dagitty('
dag {
  MoU       [exposure, pos="4,-0.5"]
  Deaths    [outcome,  pos="6.5,2"]
  Vulnerability      [latent,   pos="4,2"]
  Weather   [pos="0,2"]
  FE        [adjusted, pos="1.5,3.5"]
  SAR       [pos="3,0.5"]
  Boats     [pos="5,0.5"]
  Crossings [pos="1.5,1.1"]

  MoU -> SAR
  MoU -> Boats
  MoU -> Crossings
  SAR -> Crossings
  SAR -> Vulnerability
  Boats -> Crossings
  Boats -> Vulnerability
  Weather -> Crossings
  Weather -> Vulnerability
  Vulnerability -> Deaths
  Crossings -> Deaths
  FE -> Weather
  FE -> Crossings
  FE -> Vulnerability
  FE -> Deaths
}
')
```

```{r dag-plot, fig.height=5}
library(ggdag)
library(ggplot2)

ggdag_status(dag, text = FALSE) +
  geom_label(aes(label = name),
             fill = "grey93", colour = "black",
             size = 3, fontface = "bold",
             label.padding = unit(0.2, "lines"),
             label.size = 0, na.rm = TRUE) +
  theme_dag() +
  guides(color = guide_legend(title = "Node status")) +
  ggtitle("DAG: MoU → Deaths (Vulnerability unobserved)")
```

::: {.callout-note}
### Notes on edge directionality and omitted edges

**SAR ↔ Crossings (bidirectional in theory, unidirectional in the
DAG).** The DAG includes SAR → Crossings, representing the post-MoU
deterrence channel: reduced SAR capacity reduces crossings
(Rodriguez Sanchez et al. 2023, Model C; Hoffmann Pham & Komiyama
2024, ECM). However, the reverse — Crossings → SAR — is also
plausible, especially in the pre-MoU period: the migration crisis
drove the expansion of SAR operations (Mare Nostrum was launched in
response to the October 2013 Lampedusa shipwreck; NGO SAR expanded
as crossings surged). Rodriguez Sanchez et al. (2023) are explicit
that SAR is partly a consequence of crossing attempts, not just a
cause. In the post-MoU period that our model targets, the dominant
direction is SAR → Crossings (policy-driven SAR withdrawal reduces
crossings). The DAG represents this post-treatment causal direction.

**MoU → Crossings (direct deterrence channel).** The MoU reduced
crossings not only through SAR withdrawal but also through direct
deterrence mechanisms outside the SAR/Boats pathway: tightening of
Libya's southern border, detention of migrants in Libya, and
signaling effects that discouraged departure (Zambiasi & Albarosa
2025). This edge is included in the DAG for completeness. Since
MoU has no parents, it does not create backdoor paths.

**SAR → Boats (not included).** Deiana et al. (2024) show that
extensive SAR *enables* inflatable boat use — their entire
identification rests on SAR making unsafe boats viable. This
suggests SAR → Boats may exist as a causal edge. We do not include
it in the main DAG because (a) the MoU independently affected both
SAR and Boats through distinct policy channels (SAR externalization
vs. smuggler market disruption), and (b) adding SAR → Boats would
not change identification (all paths remain front-door from MoU,
which is parentless). However, the mechanism is real and should be
acknowledged: part of the boat-type shift documented by Deiana et
al. and Hoffmann Pham & Komiyama (2024) is mediated by SAR
availability rather than being a direct MoU effect.

**MoU → Vulnerability directly (not included).** The MoU shifted migrants
to more dangerous conditions in Libyan detention and exploitation
(documented by Zambiasi & Albarosa 2025). If pre-crossing
vulnerability feeds into at-sea crossing vulnerability, a direct
MoU → Vulnerability edge could exist. We omit it from the main DAG because
(a) Vulnerability as defined here captures *at-sea crossing vulnerability*
specifically, not pre-departure conditions, and (b) even if
included, it would not threaten identification since MoU has no
parents. This channel is relevant for the broader discussion of
MoU consequences but is outside the scope of the Weather × Post
identification strategy.
:::

::: {.callout-important}
### Detection and sample selection

The DAG models the data generating process for **all** incidents
on the CMR. However, IOM MMP records only incidents where at
least one person died or went missing ($n = 1{,}388$ CMR
incidents; every record has dead + missing $\geq 1$). The data
also records survivors when known (577/1,388 incidents).

**Incident-formation selection.** Distress events where SAR
rescued everyone (0 dead, 0 missing) do not appear in MMP. This
is inherent to the data source. Pre-MoU, when SAR capacity was
high, incidents still recorded deaths alongside many survivors
(mean survivors = 243.7 pre-MoU vs.\ 61.1 post-MoU, among
incidents with recorded survivors), indicating that SAR
responded but could not prevent all fatalities. The pre-MoU
weather--mortality gradient is estimated from this variation in
severity across weather conditions. The key question is whether
fully-rescued events (0 dead, 0 missing) are disproportionately
concentrated in bad weather pre-MoU --- if so, the pre-MoU
gradient is attenuated. This is difficult to assess directly but
is partially addressed by the weather common support diagnostic
(`estimands.qmd` Section 2.3).

**Detection selection (genuine threat).** Post-MoU, the
detection environment changed: NGO SAR vessels --- which
witnessed and reported many incidents --- were restricted
(Cusumano & Villa 2019), and the Libyan Coast Guard does not
systematically report interceptions-gone-wrong. If detection
probability changed differentially with weather post-MoU
(e.g., NGO vessels that operated disproportionately in bad
weather withdrew), the observed Weather × Post interaction
conflates true danger amplification with detection-driven
composition change.

Detection bias is not modeled in the DAG because it operates on
the *observation process*, not on the causal data generating
process for outcomes. It is addressed by the parameterized
sensitivity analysis in `estimands.qmd` Section 3.6, Exercise 3,
which models detection probability as a function of weather and
period, and asks how much detection must have worsened to nullify
the result.
:::

## 2. All causal paths

### MoU → Deaths

```{r paths-mou}
p_mou <- paths(dag, from = "MoU", to = "Deaths")
data.frame(
  status = ifelse(p_mou$open, "OPEN", "closed"),
  path   = p_mou$paths
)
```

### Weather → Deaths

```{r paths-weather}
p_wx <- paths(dag, from = "Weather", to = "Deaths")
data.frame(
  status = ifelse(p_wx$open, "OPEN", "closed"),
  path   = p_wx$paths
)
```

## 3. Adjustment sets

### Total effect: MoU → Deaths

```{r adj-mou}
adj_mou <- adjustmentSets(dag, exposure = "MoU", outcome = "Deaths",
                           type = "minimal")
if (length(adj_mou) == 0) {
  cat("No adjustment needed — no backdoor paths from MoU to Deaths.\n")
  cat("MoU is exogenous: it has no parents in the DAG.")
} else {
  print(adj_mou)
}
```

MoU has no parents → no backdoor paths → the total effect is
identified without any conditioning.

### Total effect: Weather → Deaths

```{r adj-weather}
adjustmentSets(dag, exposure = "Weather", outcome = "Deaths",
               type = "minimal")
```

Must condition on **FE** to block the backdoor: Deaths ← FE → Weather.

### Canonical adjustment sets

```{r adj-canonical}
cat("MoU → Deaths:\n")
print(adjustmentSets(dag, exposure = "MoU", outcome = "Deaths",
                      type = "canonical"))

cat("\nWeather → Deaths:\n")
print(adjustmentSets(dag, exposure = "Weather", outcome = "Deaths",
                      type = "canonical"))
```

## 4. Checking specific conditioning strategies

```{r check-adj}
results <- data.frame(
  estimand = c(rep("MoU → Deaths", 3), rep("Weather → Deaths", 3)),
  conditioning_set = c(
    "{FE}", "{FE, Weather}", "{} (empty)",
    "{FE}", "{FE, MoU}", "{} (empty)"
  ),
  valid = c(
    isAdjustmentSet(dag, "MoU", "Deaths", c("FE")),
    isAdjustmentSet(dag, "MoU", "Deaths", c("FE", "Weather")),
    isAdjustmentSet(dag, "MoU", "Deaths", c()),
    isAdjustmentSet(dag, "Weather", "Deaths", c("FE")),
    isAdjustmentSet(dag, "Weather", "Deaths", c("FE", "MoU")),
    isAdjustmentSet(dag, "Weather", "Deaths", c())
  )
)
knitr::kable(results)
```

::: {.callout-note}
For the Weather × Post interaction model:

- The primary (daily panel) model conditions on **week-by-year FE**
  (= FE node); the complementary event model uses
  grid + year + month FE ✓
- Weather is exogenous to the MoU ✓
- The interaction $\beta_3$ captures the **change** in Weather's total
  effect on Deaths across periods
:::

## 5. Collider bias: conditioning on Crossings (Kronmal problem)

This is the DAG representation of **Kronmal (1993)**. When we use
mortality **rate** = Deaths/Crossings as the outcome, we implicitly
condition on Crossings.

Crossings is a **collider** between multiple parent nodes:

- Weather → Crossings ← SAR
- Weather → Crossings ← Boats

Conditioning on a collider **opens** spurious non-causal paths.

```{r collider-check}
cat("Is {FE, Crossings} valid for Weather → Deaths?\n")
cat(isAdjustmentSet(dag, "Weather", "Deaths", c("FE", "Crossings")))
```

### What paths open when we condition on Crossings?

```{r collider-paths}
dag_cond_cross <- dag
adjustedNodes(dag_cond_cross) <- c("FE", "Crossings")
p_cond <- paths(dag_cond_cross, from = "Weather", to = "Deaths")
data.frame(
  status = ifelse(p_cond$open, "OPEN", "closed"),
  path   = p_cond$paths
)
```

::: {.callout-warning}
**Kronmal problem in DAG terms:** conditioning on Crossings opens
collider paths through SAR and Boats:

- Weather → \[Crossings\] ← SAR → Vulnerability → Deaths
- Weather → \[Crossings\] ← Boats → Vulnerability → Deaths

The mortality **rate** inherits denominator-driven bias because
Crossings sits as a collider between Weather and SAR/Boats, which
also causally affect Deaths through Vulnerability.
:::

## 6. Can we fix collider bias by also conditioning on SAR + Boats?

```{r fix-collider}
cat("Is {FE, Crossings, SAR, Boats} valid for Weather → Deaths?\n")
cat(isAdjustmentSet(dag, "Weather", "Deaths",
                    c("FE", "Crossings", "SAR", "Boats")))
```

```{r fix-collider-paths}
dag_cond_fix <- dag
adjustedNodes(dag_cond_fix) <- c("FE", "Crossings", "SAR", "Boats")
p_fix <- paths(dag_cond_fix, from = "Weather", to = "Deaths")
data.frame(
  status = ifelse(p_fix$open, "OPEN", "closed"),
  path   = p_fix$paths
)
```

::: {.callout-note}
If SAR and Boats were fully observed, conditioning on
{FE, Crossings, SAR, Boats} would block the collider paths and
allow using the mortality rate. But SAR and Boats are only
**partially observed** (and Vulnerability remains latent), so this fix is
not practically available.
:::

## 7. Implied conditional independencies

These are **testable** predictions of the DAG. If any of these fail
empirically, the DAG structure may be wrong.

```{r cond-indep}
ci <- impliedConditionalIndependencies(dag)
for (s in ci) cat(format(s), "\n")
```

## 8. Instrumental variables

```{r iv}
cat("For MoU → Deaths:\n")
iv_mou <- instrumentalVariables(dag, exposure = "MoU", outcome = "Deaths")
if (length(iv_mou) == 0) {
  cat("  None needed (MoU is exogenous, no backdoor paths)\n")
} else {
  print(iv_mou)
}

cat("\nFor Weather → Deaths:\n")
iv_wx <- instrumentalVariables(dag, exposure = "Weather",
                                outcome = "Deaths")
if (length(iv_wx) == 0) {
  cat("  None found\n")
} else {
  print(iv_wx)
}
```

## 9. Sensitivity: adding Crossings → Vulnerability

Overcrowding from high volume could increase vulnerability. This adds
a new causal chain: Weather → Crossings → Vulnerability → Deaths. Since FE →
SAR and FE → Boats are absorbed by conditioning on FE, we only test
this edge.

```{r sensitivity-crossings-Vulnerability}
dag_ext <- dagitty('
dag {
  MoU       [exposure, pos="4.5,-0.5"]
  Deaths    [outcome,  pos="6.5,2"]
  Vulnerability      [latent,   pos="4,2"]
  Weather   [pos="0,2"]
  FE        [adjusted, pos="2.5,4"]
  SAR       [pos="2.5,0.5"]
  Boats     [pos="5,0.5"]
  Crossings [pos="0.5,3"]

  MoU -> SAR
  MoU -> Boats
  MoU -> Crossings
  SAR -> Crossings
  SAR -> Vulnerability
  Boats -> Crossings
  Boats -> Vulnerability
  Weather -> Crossings
  Weather -> Vulnerability
  Vulnerability -> Deaths
  Crossings -> Deaths
  Crossings -> Vulnerability
  FE -> Weather
  FE -> Crossings
  FE -> Vulnerability
  FE -> Deaths
}
')

cat("With Crossings → Vulnerability added:\n\n")

cat("Adjustment sets for Weather → Deaths:\n")
print(adjustmentSets(dag_ext, exposure = "Weather", outcome = "Deaths",
                      type = "minimal"))

cat("\nAdjustment sets for MoU → Deaths:\n")
adj_ext <- adjustmentSets(dag_ext, exposure = "MoU", outcome = "Deaths",
                           type = "minimal")
if (length(adj_ext) == 0) {
  cat("  Still no adjustment needed (MoU still exogenous)\n")
} else {
  print(adj_ext)
}
```

```{r sensitivity-collider-worse}
cat("Is {FE, Crossings} still invalid for Weather → Deaths?\n")
cat(isAdjustmentSet(dag_ext, "Weather", "Deaths", c("FE", "Crossings")))

cat("\n\nPaths Weather → Deaths | {FE, Crossings} (extended DAG):\n")
dag_ext_cond <- dag_ext
adjustedNodes(dag_ext_cond) <- c("FE", "Crossings")
p_ext <- paths(dag_ext_cond, from = "Weather", to = "Deaths")
data.frame(
  status = ifelse(p_ext$open, "OPEN", "closed"),
  path   = p_ext$paths
)
```

::: {.callout-important}
Adding Crossings → Vulnerability does **not** threaten identification (still
a front-door causal path, no new backdoor). But it makes the
**Kronmal problem worse**: conditioning on Crossings now also blocks
the real causal path Weather → Crossings → Vulnerability → Deaths, on top of
opening the collider bias paths. The mortality rate becomes an even
more distorted estimand.
:::

## 10. Estimand decomposition and model hierarchy

Per-crossing mortality can be decomposed as:

$$\frac{\text{deaths}_t}{\text{crossings}_t} = \underbrace{\frac{\text{incidents}_t}{\text{crossings}_t}}_{\text{frequency}} \times \underbrace{\frac{\text{deaths}_t}{\text{incidents}_t}}_{\text{severity}}$$

This is the **preferred theoretical estimand**
$E[\text{deaths} \mid \text{crossing}]$: it isolates per-crossing
danger from volume effects. However, it is **infeasible** as a
primary analysis for two reasons:

1. **Crossings data are monthly only** (interception data from
   LCG/TCG is monthly), so the per-crossing rate cannot be computed
   at event level.
2. **Crossings is a post-treatment variable** affected by MoU. Per
   the DAG (Section 5), conditioning on Crossings opens collider
   paths through SAR and Boats. This is only fixable by also
   conditioning on SAR and Boats (Section 6), which are not fully
   observed (Boats data requested from Frontex; SAR data partially
   available).

### Alternative: monthly count decomposition

An alternative decomposition avoids the crossings denominator:

$$\text{deaths}_t = \underbrace{\text{incidents}_t}_{\text{frequency}} \times \underbrace{E[\text{deaths} \mid \text{incident}]_t}_{\text{severity}}$$

This is also an accounting identity. It does **not** require
crossings — no collider opens. However, the frequency term
($\text{incidents}_t$) is a count, not a probability per crossing.
It conflates two distinct channels:

- **Volume**: more crossings → more incidents (regardless of
  per-crossing danger)
- **Per-crossing danger**: each crossing more likely to become an
  incident

Without crossings data in the denominator, these two channels
cannot be separated within a single model.

### Model hierarchy

| Model | Estimand | Data level | Crossings needed? | What $\beta_3$ captures |
|-------|----------|------------|-------------------|------------------------|
| **Daily panel** (primary) | $E[\text{deaths}_t \mid \text{weather}_t, \text{post}_t]$ | Daily ($n = 3{,}218$) | No | Within-week change in weather--mortality gradient. Conflates extensive (more incidents) + intensive (more deaths/incident) margins |
| **Event-level** (complementary) | $E[\text{deaths}_i \mid \text{incident}_i, \text{weather}_i, \text{post}_i]$ | Event ($n \approx 1{,}200$) | No | Per-incident severity. Coarser FE (grid + yr + mo). Gives **opposite sign** to daily panel |
| **Monthly death counts** | $E[\text{deaths}_t \mid \text{weather}_t, \text{post}_t]$ | Monthly | No | Did more people **die in total** in bad-weather months after MoU? Mixes volume + danger |
| **Crossings diagnostic** | $E[\text{crossings}_t \mid \text{weather}_t, \text{post}_t]$ | Monthly | As outcome, not conditioning | Did weather's effect on **crossing volume** change after MoU? (volume channel assessment; see Section 12) |
| **D** (exploratory rate) | $\text{deaths}_t / \text{crossings}_t \sim \text{weather} \times \text{post}$ | Monthly | Yes → collider bias | Per-crossing danger, but collider bias from SAR/Boats (see Section 13) |
| **C** (aspirational) | As D, conditioning on {FE, Crossings, SAR, Boats} | Monthly | Yes, with SAR + Boats fix | Per-crossing danger, collider-free — requires SAR and Boats data |

The **daily panel** is the primary analysis: week-by-year FE
provide tight identification from day-to-day weather variation
within the same week. The event-level model is complementary,
isolating the per-incident severity margin with coarser FE. The
crossings diagnostic (Section 12) helps disentangle volume from
danger. Model D is exploratory with acknowledged collider bias.
Model C is theoretically preferred but requires data not yet
available.

## 12. Volume channel diagnostic: Crossings ~ Weather × Post

To partially disentangle the volume and danger channels in the
monthly death count model, we can model crossings as an **outcome**
in a separate regression:

$$\text{crossings}_t \sim f(\text{weather}_t \times \text{Post}_t + \text{FE})$$

This models crossings as an **outcome**, not as a conditioning
variable — so no collider problem arises.

### DAG validity

```{r adj-weather-crossings}
cat("Adjustment sets for Weather → Crossings:\n")
print(adjustmentSets(dag, exposure = "Weather",
                      outcome = "Crossings",
                      type = "minimal"))

cat("\nIs {FE} valid for Weather → Crossings?\n")
cat(isAdjustmentSet(dag, "Weather", "Crossings", c("FE")))
```

```{r paths-weather-crossings}
p_wx_cross <- paths(dag, from = "Weather", to = "Crossings")
data.frame(
  status = ifelse(p_wx_cross$open, "OPEN", "closed"),
  path   = p_wx_cross$paths
)
```

The only paths from Weather to Crossings are:

- Weather → Crossings (direct causal edge — the one we want)
- Weather ← FE → Crossings (backdoor through FE — blocked by
  conditioning on FE)

No other paths exist because: (a) Weather's only parent in the DAG
is FE, and (b) the path Weather → Vulnerability → Deaths ← Crossings has a
collider at Deaths, which is blocked unless we condition on Deaths
(we do not in this regression).

Conditioning on **{FE}** is sufficient. The regression is **valid
per the DAG**.

::: {.callout-note}
Crossings is only a problem when it appears in the **conditioning
set** of a regression targeting Weather → Deaths (collider). When
Crossings is the **outcome**, there is no collider issue — we are
modelling it, not conditioning on it. SAR and Boats do not need to
be included because they do not cause Weather (they are not
confounders of the Weather → Crossings relationship).
:::

### Interpretation

$\beta_3$ in this regression captures whether the MoU changed how
weather affects crossing volume. Pre-MoU, bad weather strongly
deters crossings because SAR makes waiting costless (Deiana et al.
2024). Post-MoU, smugglers may be less responsive to weather
conditions.

This serves as a **diagnostic**: if weather's effect on crossings
did not change much post-MoU, then $\beta_3$ in the monthly death
count model is primarily about the **danger channel**
(Weather → Vulnerability → Deaths), not the volume channel
(Weather → Crossings → Deaths). This is a mediation-style argument
without formal mediation analysis, but it is informative and avoids
the collider problem entirely.

## 13. Rate regression as exploratory analysis (Model D)

The theoretical estimand
$E[\text{deaths} \mid \text{crossing}]$ is the most substantively
meaningful measure of route danger. It can be operationalized as a
rate regression:

$$\log(\text{death rate}_t + c) \sim \beta_1 \text{weather}_t + \beta_2 \text{Post}_t + \beta_3 (\text{weather}_t \times \text{Post}_t) + \lambda_m$$

or equivalently as a count model with offset:

$$\text{deaths}_t \sim \text{NegBin}\!\bigl(\text{weather} \times \text{Post} + \text{offset}(\log \text{crossings}_t) + \text{month FE}\bigr)$$

Both formulations condition on Crossings, which opens the collider
paths documented in Section 5.

### Why $\beta_3$ is partially protected by differencing

The Kronmal-induced bias from the denominator exists in **both**
the pre- and post-MoU periods (the Weather → Crossings relationship
is present throughout). The interaction $\beta_3$ captures the
**change** in the weather–mortality relationship across periods. If
the denominator-driven bias were constant across periods, it would
difference out.

Moreover, the death rate has clear substantive meaning: fewer
crossings with equal deaths implies higher per-crossing danger —
which is the real phenomenon.

### Why this argument is incomplete

The collider bias is **not** constant across periods. The spurious
paths opened by conditioning on Crossings are:

- Weather → \[Crossings\] ← **SAR** → Vulnerability → Deaths
- Weather → \[Crossings\] ← **Boats** → Vulnerability → Deaths

The MoU changed both SAR (withdrawal) and Boats (shift to
inflatables). Pre-MoU, SAR is high and Boats are predominantly
wooden → one collider bias magnitude. Post-MoU, SAR is low and
Boats are inflatable → a different magnitude. Therefore, $\beta_3$
picks up both the **true change** in weather-lethality and the
**change in collider bias** across periods. The direction of the
additional bias is ambiguous without knowing the exact structural
parameters.

### How to use Model D

Model D remains informative as an **exploratory/robustness**
analysis:

- Run with acknowledged collider bias
- Compare $\beta_3$ with and without crossings in the denominator
  (i.e., compare with the monthly death count model) to assess the
  empirical magnitude of the collider distortion
- If Frontex boat-type data become available, add Boats as a
  control (partially closing the collider)
- If SAR proxy data are available (e.g., NGO fleet presence), add
  SAR as well (approaching the Model C ideal of Section 6)
- Report alongside the daily panel (primary) and monthly counts
  (complementary) to show robustness

::: {.callout-important}
Model D is **not** the primary analysis. It is exploratory evidence
that the per-crossing danger signal exists, reported with
transparent discussion of the collider bias and its likely
direction.
:::

## 14. Primary specification: daily panel NegBin

### Design and specification

The primary model is a daily panel negative binomial regression
following the Deiana et al. (2024) design
(`03c_daily_panel_model.R`):

$$\log \mu_t = \lambda_{wy} + \beta_1 \, \text{SWH}_t +
\beta_3 \, (\text{SWH}_t \times \text{Post}_t) + \gamma \,
\text{Wind}_t$$

where:

- $t$ indexes days ($n = 3{,}218$, from 2014 to 2025)
- $\mu_t$ = expected daily count of dead + missing on the CMR
- $\lambda_{wy}$ = week-by-year fixed effects ($\approx$460 cells)
- $\text{SWH}_t$ = daily spatial mean significant wave height over
  the core CMR channel [11°E--15°E, 32°N--36°N]
- $\text{Post}_t$ = 1 for dates $\geq$ 2017-02-01 (main effect
  absorbed by week-year FE)
- $\text{Wind}_t$ = daily spatial mean wind speed (control)

### Identification

Week-by-year FE absorb all confounders that vary at weekly or
slower frequency: seasonality, annual trends, policy shifts, and
slow-moving changes in SAR capacity, boat types, and smuggling
networks. $\beta_3$ is identified from **day-to-day weather
variation within the same week**, comparing how this variation
maps to deaths pre- vs.\ post-MoU.

This is substantially tighter identification than the
complementary event-level model (`03d_event_model_revised.R`),
which uses grid $\times$ year $\times$ month FE and relies on
within-month cross-sectional variation.

### Why negative binomial?

The daily death count has extreme overdispersion: most days
record zero deaths ($\approx$95% zeros), but days with
incidents can record from 1 to several hundred dead/missing.
The variance/mean ratio is approximately 132. The Poisson
assumption (mean = variance) is severely violated.

The negative binomial adds a dispersion parameter $\theta$
that accommodates this overdispersion. In the primary
specification, $\hat{\theta} \approx 0.158$, indicating very
high overdispersion (small $\theta$ = more dispersion).

**Alternatives considered:**

| Model | Issue for this setting |
|-------|----------------------|
| Poisson | Mean = variance severely violated (var/mean $\approx$ 132); would understate standard errors |
| Quasi-Poisson | Robust to overdispersion but less efficient; does not model dispersion structure |
| Zero-inflated NB | High zero fraction could motivate ZINB, but NegBin already handles excess zeros through the dispersion parameter; ZINB adds complexity without clear gain |
| OLS on $\log(\text{deaths} + 1)$ | Arbitrary constant, loses count structure, inconsistent under heteroskedasticity |

### Weather variables

The primary weather variable is `swh_core`: the daily spatial mean
of ERA5 significant wave height over the core CMR channel
[11°E--15°E, 32°N--36°N], where approximately 80% of incidents
occur. This captures the conditions experienced by boats on the
Libya--Lampedusa corridor.

Wind speed (`wind_core`) is included as a control. The interaction
uses SWH because it is the most physically direct measure of sea
danger: SWH determines wave overtopping, capsize risk, and the
Rayleigh-distributed maximum wave height (max wave $\approx 2
\times$ SWH).

Robustness variants (all in `03c_daily_panel_model.R`):

- **Lag-1 timing** (departure-day weather): same sign, attenuated
  ($\hat{\beta}_3 = +0.57$, $p = 0.27$)
- **Wind $\times$ Post**: marginal ($\hat{\beta}_3 = +0.15$,
  $p = 0.10$)
- **Gust $\times$ Post**: marginal ($\hat{\beta}_3 = +0.13$,
  $p = 0.06$)
- **SWH alone** (no wind control): similar to primary
  ($\hat{\beta}_3 = +0.96$, $p = 0.02$)

### Standard errors

The primary specification uses heteroskedasticity-robust SEs
(`vcov = "hetero"` in fixest). Under this choice,
$\hat{\beta}_3 = +0.94$ (SE = 0.42, $p = 0.026$).

Clustering by week-year roughly doubles the SE:
$\hat{\beta}_3 = +0.94$ (SE = 0.74, $p = 0.21$). The difference
reflects within-week-year serial correlation in daily death
counts --- distress events often cluster over consecutive days.
The choice between robust and clustered SEs is consequential for
inference.

### Interpreting $\beta_3$

$\beta_3$ is the **change in the SWH--mortality gradient**
post-MoU (see `estimands.qmd` Section 1.1). On the
multiplicative scale:

> For each additional 1m of SWH (within the same week), the
> expected daily death count on the CMR post-MoU is
> $\exp(\hat{\beta}_3) = \exp(0.94) = 2.55$ times what it would
> be based on the pre-MoU weather--mortality gradient.

**Scaling note (Deiana et al. 2024):** small changes in SWH
translate to large changes in maximum wave height encountered. The
Rayleigh distribution implies max wave $\approx 2 \times$ SWH.
Marginal effects should be evaluated at meaningful SWH values
(median, 75th percentile).

::: {.callout-warning}
### Diagnostic caveat

The placebo tests and gradient stability analysis (Section 15)
show that $\hat{\beta}_3 = 0.94$ **does not uniquely reflect the
MoU**. Many alternative break dates produce larger estimates, and
the gradient oscillates rather than shifting discretely. See
Section 15 for full diagnostic results.
:::

### Complementary: event-level model

The event-level model (`03d_event_model_revised.R`) uses
individual incidents as the unit of observation ($n \approx$
1,200--1,380) with grid $\times$ year $\times$ month FE. It
isolates the **per-incident severity margin** but requires coarser
FE (too few incidents per week-year cell for the Deiana-style
design).

The event-level $\hat{\beta}_3$ is **opposite in sign**: $-0.20$
($p = 0.32$) for incident-location weather, $-0.41$ ($p = 0.12$)
for core-geography weather. This discordance is discussed in the
thesis and in Section 15.

## 15. Empirical diagnostics: threats to identification

The identification strategy assumes that the weather--mortality
gradient was stable before the MoU, and that the MoU induced a
discrete, permanent shift. Diagnostics in `03b_placebo_stability.R`
test both assumptions using the primary daily panel specification.

### Placebo treatment dates

Re-estimating $\beta_3$ at 15 candidate break dates (quarterly,
2015-Q1 to 2019-Q1) reveals that the actual MoU date (Feb 2017)
does **not** stand out:

- Point estimates for 2015--2016 placebo dates are *larger* than
  the MoU estimate: $\hat{\beta}_3 \approx 1.3$--$1.8$ vs.\
  $0.94$ at the MoU date
- Where SEs are computable (2016-Q2 onward), the significant
  positive interaction starts well before the MoU
  ($\hat{\beta}_3 = 1.45$, $p = 0.005$ at 2016-Q4)
- From 2016 to 2019, $\hat{\beta}_3$ declines monotonically ---
  the MoU date sits on this declining trajectory, not as a
  discrete spike

This pattern is consistent with a **trending gradient** rather
than a discrete break at the MoU date.

### Gradient instability

Pre-period year-by-year SWH slopes (separate Poisson regressions
on event-days) show substantial variation across pre-MoU years
(2014--2016). Rolling 2-year window estimates over the full
sample (2014--2025) reveal the gradient oscillates: negative
2015--2016, positive 2017--2019, negative again 2020--2021.

This violates a core assumption of the interaction design: that
$g_{\text{pre}}(w)$ is a stable baseline against which
$g_{\text{post}}(w)$ can be compared. If the gradient was already
shifting before the MoU, $\hat{\beta}_3$ captures the MoU effect
**plus** any pre-existing trend in gradient dynamics.

### Event model discordance

The complementary event-level model
(`03d_event_model_revised.R`) produces $\hat{\beta}_3$ **opposite
in sign** to the daily panel:

| Model | $\hat{\beta}_3$ | SE | $p$ | IRR |
|-------|-----------------|------|-------|------|
| Daily panel (primary) | $+0.94$ | 0.42 | 0.026 | 2.55 |
| Event-level (incident-location SWH) | $-0.20$ | 0.20 | 0.32 | 0.82 |
| Event-level (core-geography SWH) | $-0.41$ | 0.26 | 0.12 | 0.66 |

The discordance likely reflects: (a) different FE granularity
(week-year vs.\ grid + year + month), (b) the daily panel
conflating the extensive margin (more incidents on bad-weather
days) with the intensive margin (more deaths per incident), and
(c) composition effects in which incidents enter the MMP sample.

### Implications for the thesis

These diagnostics undermine a **causal MoU interpretation** of
$\hat{\beta}_3$. The thesis documents the gradient instability as
the primary empirical finding and discusses its implications:

1. **The MoU date is not special** in the placebo test --- many
   dates produce similar or larger estimates
2. **The gradient oscillates** over the full period, inconsistent
   with a permanent structural break
3. **The daily panel and event models disagree** on the sign of
   $\beta_3$, suggesting the result is sensitive to specification
4. **Clustered SEs** (accounting for within-week correlation)
   raise $p$ to 0.21, further weakening the statistical case

The identification strategy from the DAG (Sections 1--9) remains
**logically valid**: if the assumptions hold, $\beta_3$ identifies
the MoU's effect on the weather--mortality gradient. The problem
is **empirical**: the data do not support the assumption of a
stable pre-period gradient that shifted discretely at the MoU.

## 16. Adjustment set visualization

```{r dag-adj-plot, fig.height=5}
ggdag_adjustment_set(dag, text = FALSE) +
  geom_label(aes(label = name),
             fill = "grey93", colour = "black",
             size = 3, fontface = "bold",
             label.padding = unit(0.2, "lines"),
             label.size = 0, na.rm = TRUE) +
  theme_dag() +
  ggtitle("Required adjustments: Weather → Deaths")
```

## 17. Summary

| Finding | Implication |
|---------|-------------|
| **MoU → Deaths**: no backdoor paths | Total effect identified without conditioning (MoU is exogenous) |
| **Weather → Deaths**: one backdoor through FE | Conditioning on FE (week-year in daily panel) is **sufficient** |
| **Conditioning on Crossings** (= using rate) | Opens collider paths through SAR/Boats → **Kronmal bias** |
| **Vulnerability is latent** | Weather × Post interaction identifies MoU's effect on the Weather → Vulnerability → Deaths pathway **without** observing Vulnerability |
| **Exclusion restriction** | No direct Weather × Post → Deaths path outside Vulnerability must hold |
| **Per-crossing decomposition infeasible** | Crossings only monthly + post-treatment → collider. Daily panel (severity + volume) is primary; event model complementary |
| **Crossings ~ Weather × Post is valid** | {FE} sufficient; no collider because Crossings is outcome, not conditioning set. Diagnostic for volume channel |
| **Rate regression (Model D)** | $\beta_3$ partially protected by differencing, but collider bias changes across periods (SAR/Boats shift). Exploratory only |
| **Empirical diagnostics** | Placebo tests show MoU date is not special; gradient oscillates over full period; event model gives opposite sign to daily panel. Identification strategy is logically valid but empirically undermined (Section 15) |

### Data needs

| Variable | Status | Source |
|----------|--------|--------|
| Weather (SWH, wind, SST) | **Observed** | ERA5 |
| Deaths | **Observed** | IOM MMP |
| FE (date, grid) | **Observed** | Date/coordinates |
| SAR | Mostly unobserved (low quality data) | NGO fleet data |
| Boats | Partially observed | Frontex JORA |
| Crossings | Observed — **do not condition on** in primary model; use as **outcome** in diagnostic (Section 12); use as **denominator** only in exploratory Model D (Section 13) | IOM/UNHCR monthly |
| Vulnerability | **Unobserved** | No single data source |

### Potential missing edges

| Edge | Plausible? | Impact on identification |
|------|-----------|------------------------|
| FE → SAR | Yes (seasonal SAR deployment) | **None** — FE is already conditioned on, so any path through FE is blocked |
| FE → Boats | Yes (seasonal boat availability) | **None** — same reason, absorbed by FE conditioning |
| Crossings → Vulnerability | Yes (overcrowding from high volume) | Adds a new **causal** front-door path (Weather → Crossings → Vulnerability → Deaths) but does not create backdoor bias. Makes conditioning on Crossings (= using the rate) **even more problematic**: it would block this additional real causal variation. Tested in Section 9 |
| Crossings → SAR | Yes in pre-MoU period (migration crisis drove SAR expansion) | Not included because the DAG represents the post-MoU causal structure. In this period, the dominant direction is SAR → Crossings (policy-driven deterrence). See discussion in Section 1 callout |
| SAR → Boats | Yes (Deiana et al. 2024: SAR enables inflatable boat use) | **None** — all paths remain front-door from MoU (parentless). Not included but acknowledged as a real mechanism. See discussion in Section 1 callout |
| MoU → Vulnerability directly | Possible (detention conditions, pre-departure exploitation) | **None** — MoU has no parents. Not included because Vulnerability here captures at-sea vulnerability specifically. Relevant for broader discussion of MoU consequences |
| Weather → Deaths directly | No — every mechanism (waves capsizing boats, storms preventing rescue, exposure at sea) operates through crossing vulnerability | Not applicable; exclusion restriction is defensible |
| Detection mechanism (MoU → Detection, Weather → Detection) | Yes — post-MoU: fewer NGO witnesses; bad weather: harder to observe incidents in storms | **Not a confounding problem** but a sample selection problem. Does not create backdoor bias for the total effect, but threatens $\beta_3$ if detection probability changed differentially with weather post-MoU. Addressed by parameterized sensitivity analysis (`estimands.qmd` Section 3.6, Exercise 3), not by DAG conditioning |

### Literature support for DAG edges

| Edge | Supporting literature |
|------|---------------------|
| MoU → SAR | Zambiasi & Albarosa (2025): SAR externalized to LCG; Deiana et al. (2024): Minniti Code withdrew NGO SAR; Rodriguez Sanchez et al. (2023): EU-Libya cooperation = "shift to deterrence"; Hoffmann Pham & Komiyama (2024): rescue probability falls from ~90% to ~50%; Steinhilper & Gruijters (2018): mortality rates rose when SAR scaled back |
| MoU → Boats | Deiana et al. (2024): inflatable share drops post-Minniti; Hoffmann Pham & Komiyama (2024): smugglers shift to smaller boats as interception rises; Friebel et al. (2024): migration choices sensitive to crossing costs, including boat availability |
| MoU → Crossings | Hoffmann Pham & Komiyama (2024): deterrence via rescue probability; Zambiasi & Albarosa (2025): southern border tightening, detention deterrence; Rodriguez Sanchez et al. (2023): Model C shows pushback period reduces crossings; Friebel et al. (2024): elasticity of migration intentions to smuggling distance > −3; Tafani & Riccaboni (2025): EU-Turkey deal diverted ~2,000 migrants to CMR |
| SAR → Crossings | Hoffmann Pham & Komiyama (2024): ECM, rescue probability fall 0.9→0.5 predicts ~11,200 fewer monthly crossings; Rodriguez Sanchez et al. (2023): null at aggregate monthly level but significant for pushback period; Battiston (2022): reducing rescue distance increases future departures; Cusumano & Villa (2019): no pull factor from NGO SAR; Amenta et al. (2021): SAR not a pull factor |
| SAR → Vulnerability | Deiana et al. (2024): SAR withdrawal = no rescue for distressed boats; Zambiasi & Albarosa (2025): LCG's poor SAR increases deadly event probability by 50% per 100km from Libya; Battiston (2022): rescue distance directly affects safety; Fasani et al. (2019): SAR operations related to probability of dying; Steinhilper & Gruijters (2018): mortality rates rose when SAR scaled back |
| Boats → Vulnerability | Deiana et al. (2024): inflatables ~20× more dangerous than wooden boats (θ ≈ 5-10%); Hoffmann Pham & Komiyama (2024): boat size affects survival |
| Weather → Crossings | Deiana et al. (2024): 0.10m SWH increase reduces crossings 8.9-14.6%; Camarena et al. (2020): 10% wave height increase → ~27% fewer arrivals; Cusumano & Villa (2019): weather conditions are the main determinant of departures from Libya |
| Weather → Vulnerability | Deiana et al. (2024): weather × boat-type interaction — weather more dangerous for fragile boats; Camarena et al. (2020): rougher seas raise death risk (daily level). Empirical link at monthly frequency not yet fully explored in our analysis |
| Vulnerability → Deaths | All papers: definitional — vulnerability determines whether crossings result in death. Steinhilper & Gruijters (2018): descriptive analysis of mortality risk 2010-2016 |
| Crossings → Deaths | Tafani & Riccaboni (2025): crossings coeff = 0.001 (p=0.012) on missing migrants; Amenta et al. (2021): crossing volume as determinant of deaths |
| FE → {all} | Universal: every paper conditions on time/seasonal FE. Deiana et al. (2024): week-by-year FE; Camarena et al. (2020): month + day-of-week FE; Zambiasi & Albarosa (2025): month + quadrant FE + quadrant-specific trends |

### Structural assumptions: literature support

| Assumption | Supporting literature |
|-----------|---------------------|
| MoU exogeneity (no parents) | Zambiasi & Albarosa (2025): geopolitical agreement endorsed at EU Malta summit; Hoffmann Pham & Komiyama (2024): treats LCG shift as exogenous policy change |
| Weather ⊥ MoU | By physics: atmospheric/oceanic conditions are independent of political agreements |
| Exclusion restriction (no Weather → Deaths bypassing Vulnerability) | Deiana et al. (2024): weather's effect on danger is mediated by boat vulnerability; Camarena et al. (2020): all identified mechanisms (capsizing, failed rescue, exposure) operate through crossing vulnerability |
| Vulnerability is latent | No paper fully observes crossing vulnerability. Deiana et al. (2024): observe boat type only; Hoffmann Pham & Komiyama (2024): observe boat size and passengers only; Zambiasi & Albarosa (2025): observe location relative to SAR zones only |
| Collider bias from conditioning on Crossings | Kronmal (1993): foundational analysis of ratio-outcome bias; Elwert & Winship (2014): formalization of collider bias in DAG terms; Camarena et al. (2020): their death-rate regression is an example of this problem |
| Weather × Post interaction as identification | Novel in this context. Closest precedent: Deiana et al. (2024) use weather × SAR × boat-type triple interaction for crossing behavior, not mortality directly |

## References

1. Amenta, C., Di Betta, P., & Ferrara, C. (2021). The migrant
   crisis in the Mediterranean Sea: Empirical evidence on policy
   interventions. *Socio-Economic Planning Sciences*, 75, 100880.

2. Battiston, G. (2022). Rescue on stage: Border enforcement and
   public attention in the Mediterranean Sea. Marco Fanno Working
   Paper No. 0292.

3. Camarena, K. R., Claudy, S., Wang, Y., & Wright, A. L. (2020).
   Political and environmental risks influence migration and human
   smuggling across the Mediterranean Sea. *SSRN Working Paper*.

4. Cusumano, E., & Villa, M. (2019). Sea rescue NGOs: A pull
   factor of irregular migration? *EUI Policy Brief*.

5. Deiana, C., Maheshri, V., & Mastrobuoni, G. (2024). Migrants
   at sea: Unintended consequences of search and rescue
   operations. *American Economic Journal: Economic Policy*.

6. Elwert, F., & Winship, C. (2014). Endogenous selection bias:
   The problem of conditioning on a collider variable. *Annual
   Review of Sociology*, 40, 31--53.

7. Fasani, F., Frattini, T., & Weisser, R. (2019). The
   relationship between SAR operations and the probability of
   dying in the Central Mediterranean, 2012--2017. *Working
   Paper*.

8. Friebel, G., Manchin, M., Mendola, M., & Prarolo, G. (2024).
   International migration and illegal costs: Evidence from
   Africa-to-Europe smuggling routes. *Journal of International
   Economics*, 148, 103876.

9. Hoffmann Pham, K., & Komiyama, Y. (2024). Strategic choices of
   migrants and smugglers in the Central Mediterranean Sea.
   *Working Paper*.

10. Kronmal, R. A. (1993). Spurious correlation and the fallacy
    of the ratio standard revisited. *Journal of the Royal
    Statistical Society: Series A*, 156(3), 379--392.

11. Rodriguez Sanchez, A., Wucherpfennig, J., Rischke, R., &
    Iacus, S. M. (2023). Search-and-rescue in the Central
    Mediterranean Route does not induce migration: Predictive
    modeling to answer causal queries in migration research.
    *Scientific Reports*, 13, 12871.

12. Steinhilper, E., & Gruijters, R. J. (2018). A contested
    crisis: Policy narratives and empirical evidence on border
    deaths in the Mediterranean. *Sociology*, 52(3), 515--533.

13. Tafani, S., & Riccaboni, M. (2025). The impact of the
    EU--Turkey agreement on the number of lives lost at sea.
    *Humanities and Social Sciences Communications*, 12, 282.

14. Textor, J., van der Zander, B., Gilthorpe, M. S., Liskiewicz,
    M., & Ellison, G. T. (2016). Robust causal inference using
    directed acyclic graphs: The R package 'dagitty'.
    *International Journal of Epidemiology*, 45(6), 1887--1894.

15. Zambiasi, D., & Albarosa, E. (2025). Externalizing rescue
    operations at sea: The migration deal between Italy and Libya.
    *Journal of Economic Geography*, 25(1), 41--58.
